Management induced changes of soil organic carbon on global croplands

### 2020-12-14 SOCbudgetPaper/SOCbudget.Rmd

---
title: Management induced changes of soil organic carbon on global croplands 
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'communication')`"
author:
  - given_name: Kristine
    surname: Karstens
    affiliation: 1
    email: kristine.karstens@pik-potsdam.de
    corresponding: true
  - given_name: Benjamin Leon
    surname: Bodirsky
    affiliation: 1
    email: bodirsky@pik-potsdam.de
    orcid: https://orcid.org/0000-0002-8242-6712
  - given_name: Jan Philipp
    surname: Dietrich
    affiliation: 1 
    orcid: https://orcid.org/0000-0002-4309-6431
  - given_name: Marta
    surname: Dondini
    affiliation: 2
  - given_name: Jens
    surname: Heinke
    affiliation: 1
  - given_name: Matthias
    surname: Kuhnert
    affiliation: 2
  - given_name: Christoph
    surname: Müller
    affiliation: 1 
    orcid: https://orcid.org/0000-0002-9491-3550
  - given_name: Susanne
    surname: Rolinski
    affiliation: 1
  - given_name: Pete
    surname: Smith
    affiliation: 2
  - given_name: Isabelle
    surname: Weindl
    affiliation: 1
    orcid: https://orcid.org/0000-0002-7651-6930
  - given_name: Hermann
    surname: Lotze-Campen
    affiliation: 1
  - given_name: Alexander
    surname: Popp
    affiliation: 1
affiliation:
  - code: 1
    address: Potsdam Institute for Climate Impact Research (PIK), Member of the Leibniz Association, P.O. Box 60 12 03, 14412 Potsdam, Germany
  - code: 2
    address: Institute of Biological & Environmental Sciences, University of Aberdeen, Aberdeen, UK
  - code: 3
    address: Humboldt-Universität zu Berlin, Department of Agricultural Economics, Unter den Linden 6, 10099 Berlin, Germany
abstract: |
  Soil organic carbon (SOC) is one of the largest terrestrial carbon stocks on Earth. It is three times larger than the vegetation carbon pool and more than twice as big as the atmospheric carbon pool, when looking into the first meter of the Earth's soil profile. SOC has been depleted by anthropogenic land cover change and agricultural management. However, the latter has so far not been well represented in global carbon stock assessments. While SOC models often simulate detailed biochemical processes that lead to the accumulation and decay of SOC, the management decisions driving these biophysical processes are still little investigated at the global scale. Here we develop a spatial explicit data set for agricultural management on cropland, considering crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We combine it with the IPCC Tier 2 steady-state soil model to create a half-degree resolution data set of SOC stocks and SOC stock changes for the first 30 cm of mineral soils. We estimate that due to arable farming, soils have lost around 26 GtC relative to a counterfactual natural state in 1975. Yet, within the period 1975--2010 this SOC debt has been decreasing again by a net quantity of 4 Gt SOC, which can be mainly traced back to an increased input of C in crop residues due to higher crop productivity. We also find that SOC is very sensitive to management decisions such as residue returning indicating the necessity to incorporate better management data in soil model simulations.

>>>>

---
title: Management induced changes of soil organic carbon on global croplands 
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'communication')`"
author:
  - given_name: Kristine
    surname: Karstens
    affiliation: 1
    email: kristine.karstens@pik-potsdam.de
    orcid: https://orcid.org/0000-0002-5860-9789
    corresponding: true
  - given_name: Benjamin Leon
    surname: Bodirsky
    affiliation: 1
    email: bodirsky@pik-potsdam.de
    orcid: https://orcid.org/0000-0002-8242-6712
  - given_name: Jan Philipp
    surname: Dietrich
    affiliation: 1 
    orcid: https://orcid.org/0000-0002-4309-6431
  - given_name: Marta
    surname: Dondini
    affiliation: 2
  - given_name: Jens
    surname: Heinke
    affiliation: 1
  - given_name: Matthias
    surname: Kuhnert
    affiliation: 2
  - given_name: Christoph
    surname: Müller
    affiliation: 1 
    orcid: https://orcid.org/0000-0002-9491-3550
  - given_name: Susanne
    surname: Rolinski
    affiliation: 1
  - given_name: Pete
    surname: Smith
    affiliation: 2
  - given_name: Isabelle
    surname: Weindl
    affiliation: 1
    orcid: https://orcid.org/0000-0002-7651-6930
  - given_name: Hermann
    surname: Lotze-Campen
    affiliation: 1
  - given_name: Alexander
    surname: Popp
    affiliation: 1
affiliation:
  - code: 1
    address: Potsdam Institute for Climate Impact Research (PIK), Member of the Leibniz Association, P.O. Box 60 12 03, 14412 Potsdam, Germany
  - code: 2
    address: Institute of Biological & Environmental Sciences, University of Aberdeen, Aberdeen, UK
  - code: 3
    address: Humboldt-Universität zu Berlin, Department of Agricultural Economics, Unter den Linden 6, 10099 Berlin, Germany
abstract: |
  Soil organic carbon (SOC) is one of the largest terrestrial carbon stocks on Earth. The first meter of the Earths soils profile stores three times as much carbon as the vegetation and twice the amount of C in the atmosphere. SOC has been depleted by anthropogenic land cover change and agricultural management. However, the latter has so far not been well represented in global carbon stock assessments. While SOC models often simulate detailed biochemical processes that lead to the accumulation and decay of SOC, the management decisions driving these biophysical processes are still little investigated at the global scale. Here we develop a spatial explicit data set for agricultural management on cropland, considering crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We combine it with the IPCC Tier 2 steady-state soil model to create a half-degree resolution data set of SOC stocks and SOC stock changes for the first 30 cm of mineral soils. We estimate that due to arable farming, soils have lost around 26 GtC relative to a counterfactual natural state in 1975. Yet, within the period 1975--2010 this SOC debt has been decreasing again by a net quantity of 4 Gt SOC, which can be mainly traced back to an increased input of C in crop residues due to higher crop productivity. We also find that SOC is very sensitive to management decisions such as residue returning indicating the necessity to incorporate better management data in soil model simulations.

### 2020-12-14 SOCbudgetPaper/discussion.Rmd

This study shows that spatially explicit and time-variant historical agricultural management data changes estimations of states and trends of global SOC and SOC debt ($\Delta SOC$) considerably compared to the often used constant management assumptions. This result remains robust to variations of central model parameters and variations in the initialization of SOC stocks. While land cover change has depleted SOC stocks and increased the SOC debt, our analysis points out that the high increases in agricultural productivity may have even led to a net reduction of the SOC debt since 1975.

@@

The evaluation of SOC stock gains and losses, is complex and has several dimensions as climatic and anthropogenic effects overlap. If defining the SOC debt of 1975 as the baseline, and measuring land-use emissions on croplands as the difference between a potential natural state and the state under human interventions (see [@pugh_simulated_2015]), global croplands have acted as a carbon sink since 1975 according to our study. However, annual C sequesteration rates of 0.2 per mille are well below the promoted 4 per mille ([@minasny_soil_2017]), indicating that productivity gains on historic levels alone are not enough to meet ambitious climate targets.

@@

With regard to the time trend, our study shows the substantial impact of changing management factors on the development of $F_{SCF}$ as also indicated by the time trend of the SOC debt.

@@

The global estimates of the total SOC stock from this study are on the lower end compared to other modeled results or data driven estimates. SoilGrids ([@hengl_soilgrids250m_2017]) especially stands out for their high estimation, whereas all other sources (including our study) are comparabely similar. Looking at regional results in Fig. \@ref(fig:SOCreg), our estimates turn out to be in good agreement for most of the world, with the largest deviations for boreal moist and tropical moist areas. To avoid that this bias influences our results, which originates from uncertainties in the representation of natural land, we focus on SOC changes on cropland. Pristine natural vegetated areas (like permafrost and rain forests) without human land management thus drop out in our calculation of $\Delta SOC$ and $F^\mathrm{SCF}$.

>>>>

This study shows that spatially explicit and time-variant historic agricultural management data considerably alter estimates of the states and trends of SOC compared to the often used constant management assumptions. This result remains robust to variations of central model parameters and variations in the initialization of SOC stocks. While land cover change has depleted SOC stocks and increased the SOC debt, our analysis points out that the high increases in agricultural productivity may have even led to a net reduction of the SOC debt since 1975.

@@

The evaluation of SOC stock gains and losses is complex and has several dimensions as climatic and anthropogenic effects overlap. If defining the SOC debt of 1975 as the baseline, and measuring land-use emissions on croplands as the difference between a potential natural state and the state under human interventions (see [@pugh_simulated_2015]), global croplands have acted as a carbon sink since 1975 according to our study. However, annual C sequesteration rates of 0.2 per 1000 are well below the promoted 4 per 1000 ([@minasny_soil_2017]), indicating that productivity gains on historic levels alone are not enough to meet ambitious climate targets.

@@

With regard to the time trend, our study shows the substantial impact of changing management factors on the development of stock change factors as also indicated by the time trend of the SOC debt.

@@

The global estimates of the total SOC stock from this study are on the lower end compared to other modeled results or data driven estimates. SoilGrids ([@hengl_soilgrids250m_2017]) especially stands out for their high estimation, whereas all other sources (including our study) are comparabely similar. Looking at regional results in Fig. \@ref(fig:SOCreg), our estimates turn out to be in good agreement for most of the world, with the largest deviations for boreal moist and tropical moist areas. To avoid that this bias influences our results, which originates from uncertainties in the representation of natural land, we focus on SOC changes on cropland. Pristine natural vegetated areas (like permafrost and rain forests) without human land management thus drop out in our calculation of SOC debt and stock change factors.

### 2020-12-14 SOCbudgetPaper/results.Rmd

```{r captionSOCmaps, echo=FALSE}
captionSOCmaps <- "(a): Distribution of total global SOC stocks for the first 30 cm on cropland shows high carbon stocks in high yielding areas. (b)+(c): Absolute (b) and relative (c) SOC stocks changes compared to a potential natural state showing different hot spots of SOC dynamics. Whereas absolute losses $\\Delta SOC$ are often highest in temperate dry regions, relative losses $F^\\mathrm{SCF}$ are often larger in tropical moist areas. (d): The SOC debt between SOC under historic land use and potential natural vegetation is decreasing over time, meaning net SOC gains on global croplands over the the period 1975--2010. "
```

@@

We analyze the relative impact of individual management aspects by comparing the actual historic management scenario with counterfactual scenarios where individual management aspects are kept static at the 1975 values (Figure \@ref(fig:SOCscen)(a)). Without changes in management regimes, the global SOC debt on cropland would be increasing at a rate of $0.1\unit{GtC yr^{-1}}$. As shown by the constResidue scenario, changes in residue return rates dominate the management effects. Without the historic increase in residue returning, the global SOC debt would still increase at a rate of $0.06\unit{GtC yr^{-1}}$. Both the constManure and constTillage scenarios show only small deviations from the historic values (sequestration rate of $0.09\unit{GtC yr^{-1}}$ for both). The effect of no-tillage has been particuarly strong since the 1990s. 
The strong impact of almost doubling C inputs from crop residue biomass over a period of 35 years on agricultural SOC stocks is shown in Fig. \@ref(fig:SOCscen)(b).

@@

```{r SOCscen, out.width = "18cm", fig.cap = captionSOCscen, echo=FALSE, fig.pos="h"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/scenario_horiz.png")
```

@@

```{r SOCinit, out.width = "18cm", fig.cap = captionSOCinit, echo=FALSE, fig.pos="H"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/scenario_init.png")
```

>>>>

```{r captionSOCmaps, echo=FALSE}
captionSOCmaps <- "(a): Distribution of total global SOC stocks for the first 30 cm on cropland shows high carbon stocks in high yielding areas. (b)+(c): Absolute (b) and relative (c) SOC stocks changes compared to a potential natural state showing different hot spots of SOC dynamics. Whereas absolute losses $\\Delta SOC$ are often highest in temperate dry regions, relative losses $F^\\mathrm{SCF}$ are often larger in tropical moist areas. (d): $\\Delta SOC$ between SOC under historic land use and potential natural vegetation is decreasing over time, meaning net SOC gains on global croplands over the the period 1975--2010. "
```

@@

We analyze the relative impact of individual management aspects by comparing the actual historic management scenario with counterfactual scenarios where individual management aspects are kept static at the 1975 values (Figure \@ref(fig:SOCscen)(a)). Without changes in management regimes, the global $\Delta SOC$ on cropland would be increasing at a rate of $0.1\unit{GtC yr^{-1}}$. As shown by the constResidue scenario, changes in residue return rates dominate the management effects. Without the historic increase in residue returning, the global $\Delta SOC$ would still increase at a rate of $0.06\unit{GtC yr^{-1}}$. Both the constManure and constTillage scenarios show only small deviations from the historic values (sequestration rate of $0.09\unit{GtC yr^{-1}}$ for both). The effect of no-tillage has been particuarly strong since the 1990s. 
The strong impact of almost doubling C inputs from crop residue biomass over a period of 35 years on agricultural SOC stocks is shown in Fig. \@ref(fig:SOCscen)(b).

@@

```{r SOCscen, out.width="14cm", fig.cap = captionSOCscen, echo=FALSE}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/scenario_horiz.png")
```

@@

```{r SOCinit, out.width="10cm", fig.cap = captionSOCinit, echo=FALSE}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/scenario_init.png")
```
